---
title: "Homework 3"
author: "Mohammad Manzoor Hassan Mirza"
date: 'Fall 2020'
output: 
  html_document:
    toc: true
    toc_depth: 3
    theme: paper
    highlight: tango
---

##### This homework is due by **1:30PM ET on Wednesday, November 18**.  

### Loading packages

The code chunk below appears in the Rmd file, but won't be displayed in your html output.  

```{r, include = FALSE}
library(tidyverse)
library(knitr)
```


### Problem 1: A few simple plots

For this problem we'll use the `diamonds` dataset from the `ggplot2` package.

##### (a) Base R graphics

Use the `qplot` function to create a histogram of `carat` with the bars colored `steelblue`.  
```{r}
qplot(x = carat, data = diamonds, bins = 30,
      fill = I('steelblue'),
      xlab = "Carats")
```

##### (b) qplot histogram?

This time, use the `ggplot` function (not the `qplot` function!) to create a histogram of `depth`.  

```{r}
ggplot(data = diamonds, aes(x = depth)) + geom_histogram(fill = 'steelblue')
```

##### (c) qplot violin plots

Use `qplot`  to create violin plots showing how `price` varies across diamond `cut`.  Specify `fill = cut` to get all the boxplots to be coloured differently.

```{r}
qplot(x = cut, y = price, data = diamonds, geom = "violin", fill = cut)
```

**Hint**: For this exercise, it will be useful to know that `violin` is a geometry (`geom`) built into `ggplot2`, and that `qplot` can be called by specifying the following arguments:

```{r, eval = FALSE}
qplot(x, y, data, geom, fill)
```

### Problem 2: ggplot practice 

For this exercise we'll go back to the Cars93 data set in the MASS library

##### (a) size mapping, geom_point()

Define a `ggplot` object using the Cars93 data set that you can use to view `Price` on the y-axis, `MPG.highway` on the x-axis, and set the `size` mapping to be based on `Horsepower`.  

Use `geom_point()` to create a scatterplot from your `ggplot` object.

```{r}
Cars93 <- MASS::Cars93

obj.a <- ggplot (data = Cars93, aes(x = MPG.highway, y = Price, size = Horsepower))

obj.a + geom_point()
```

##### (b) colour mapping

Repeat part (a), this time also setting the `colour` mapping to be based on `Origin`.

```{r}
obj.b <- ggplot (data = Cars93, aes(x = MPG.highway, y = Price, size = Horsepower, colour = Origin))

obj.b + geom_point()
```

##### (c) changing color palette

Repeat part (b), this time using the `scale_colour_manual()` layer to specify that you want to use `cbPalette` as your color palette.

```{r}
cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")

obj.b + geom_point() + scale_colour_manual(values = cbPalette)
```


### Problem 3: More ggplot2 practice

##### (a) stat_smooth()

Repeat part 2(b), this time using `stat_smooth()` to add a layer showing the smoothed curve representing how `Price` varies with `MPG.highway`.  

```{r}
obj.b + geom_point() + stat_smooth(show.legend = FALSE)
```

##### (b) facet_grid() 

Use your ggplot object from 2(b) along with the `geom_point()` and `facet_grid` layers to create scatterplots of `Price` against `MPG.highway`, broken down by (conditioned on) `Origin`.  

```{r}
obj.b + geom_point() + facet_grid(. ~ Origin)
```

(Your code should produce a figure with two scatterplots, one for each level of `Origin`.)

##### (c) Overlaying regression lines.

Modify your solution to part (b) to also display regression lines for each scatterplot.

```{r}
obj.b + geom_point() + facet_grid(. ~ Origin) + geom_smooth(method = "lm")
```


### Problem 4: Visualizing data summaries

This problem uses the Adult dataset, which we load below.  The main variable of interest here is `high.income`, which indicates whether the individual's income was over $50K.  Anyone for whom `high.income == 1` is considered a "high earner".

```{r}
adult.data <- read_csv("http://archive.ics.uci.edu/ml/machine-learning-databases/adult/adult.data", 
                       trim_ws=TRUE, 
                       col_names=c("age", "type_employer", "fnlwgt", "education", "education_num","marital", 
                                   "occupation", "relationship", "race","sex", "capital_gain", "capital_loss", 
                                   "hr_per_week","country", "income"))

adult.data <- mutate(adult.data,
                     high.income = as.numeric(income == ">50K"))
```

##### (a) Income by education level

Use `dplyr` (tidyverse) functions to produce a summary table showing how many individuals there are in each `education` bin, and how the proportion of high earners varies across `education` levels.  Your output should be a tibble with column names: `education`, `count` and `high.earn.rate`.  

```{r}
edu_summary <- (adult.data) %>%
  group_by(education) %>%
  summarise(count = n(), 
            high.earn.rate = sum(high.income)/ count)

print (edu_summary)
```

##### (b) Constructing a bar chart

Using the `ggplot` and `geom_bar` commands along with your data summary from part **(a)** to create a bar chart showing the high earning rate on the y axis and `education` on the x axis.  Specify that the color of the bars should be determined by the number of individuals in each bin.

```{r}
plot_base <- ggplot(data = edu_summary, aes(x = education, 
                                            y = high.earn.rate, 
                                            fill = count))

plot_base + geom_bar(stat='identity') +
  ylab("High Earning Rate") + 
  xlab("Education") +
  guides(fill = guide_legend(title = "Individuals in each bin"))
```

##### (c) Improve upon your plot by following example 4 [from this guide](https://rpubs.com/sediaz/rotate_x_text) to rotate the x-axis text so that it doesn't overlap with itself.  (Any reasonable rotation is fine.  It doesn't have to be -90 degrees as illustrated in the linked example.)

```{r}
plot_base + geom_bar(stat='identity') +
  ylab("High Earning Rate") + 
  xlab("Education") +
  guides(fill = guide_legend(title = "Individuals in each bin")) +
  theme(axis.text.x=element_text(angle =- 70, vjust = 0.5))
```

##### (d)  Further improve upon your plot from part **(c)** by using `mutate()` and `factor()` to change the ordering of the `education` variable labels to something that reflects *ascending order* of educational attainment.  You can use your discretion in how to break ties for the post-high school educational levels.

```{r}

edu_summary <- edu_summary %>%
  mutate(education = factor(education,
                            levels = c("Preschool", "1st-4th", "5th-6th",
                                       "7th-8th", "9th", "10th", "11th",
                                       "12th", "HS-grad", "Some-college",
                                       "Bachelors", "Prof-school",
                                       "Masters", "Assoc-voc", "Assoc-acdm",
                                       "Doctorate")))

plot_base <- ggplot(data = edu_summary, aes(x = education, 
                                            y = high.earn.rate, 
                                            fill = count))

plot_base + geom_bar(stat='identity') +
  ylab("High Earning Rate") + 
  xlab("Education") +
  guides(fill = guide_legend(title = "Individuals in each bin")) +
  theme(axis.text.x=element_text(angle =- 70, vjust = 0.5))
```


##### (e) summary table with multiple splitting variables

Use `dplyr` functions to produce a summary table showing how the proportion of high earners varies across all combinations of the following variables: `sex`, `race`, and `marital` (marital status).  In addition to showing the proportion of high earners, your table should also show the number of individuals in each bin.  Your table should have column names: `sex`, `race`, `marital`, `count` and `high.earn.rate`.  

```{r}
var_summary <- adult.data %>%
  group_by(sex, race, marital) %>%
  summarise(count = n(), 
            high.earn.rate = sum(high.income)/ count)

print (var_summary)
```

##### (f) Nicer table output using `kable()`

Use the `kable()` function from the `knitr` library to display the table from part **(e)** in nice formatting.  You should use the `digits` argument to ensure that the values in your table are being rounded to a reasonable number of decimal places.  

```{r}
var_summary %>% 
  kable(format = "markdown", digits = 3)
```


**Hint**: You will have to specify the appropriate `format` argument in the `kable()` function.  The notes for Lecture 6 have a working example of this.

### Problem 5: Getting the right plot

##### (a) A more complex bar chart.

Using the table you created in 4(e), use ggplot graphics to construct a plot that looks like [the one at this link](http://www.andrew.cmu.edu/user/achoulde/94842/homework/target_fig.png)


**Hint** You may find it useful to use the following layers: `facet_grid`, `coord_flip` (for horizontal bar charts), `theme` (rotating x axis text) and `guides` (removing fill legend). 

```{r, fig.height = 4, fig.width = 8}
plots_base <- ggplot(data = var_summary, aes (x = marital, 
                                             y = high.earn.rate, fill = race))

plots_base + geom_bar(stat='identity') +
  ylab("Proportion earning over $50K per year") + 
  xlab("Marital Status") +
  theme(axis.text.x=element_text(angle = 90, vjust = 0.5)) + 
  facet_grid(sex ~ race) + coord_flip() + guides(fill = FALSE)
```

##### (b)  Hiding code with `echo` 

Repeat part **(a)**, but this time set the `echo` argument of the code chunk in such a way that the code is not printed, but the plot is still displayed.

```{r echo = FALSE, fig.height = 4, fig.width = 8}
plots_base + geom_bar(stat='identity') +
  ylab("Proportion earning over $50K per year") + 
  xlab("Marital Status") +
  theme(axis.text.x=element_text(angle = 90, vjust = 0.5)) + 
  facet_grid(sex ~ race) + coord_flip() + guides(fill = FALSE)
```